Thesis: draft

### 2021-02-15 draft.tex
continued writing opinion dynamics chapter + changed/added some small things here and there

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics.\\
The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.
% talk about history

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. In this chapter we will focus on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure. Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis. Some major attention will also be given to the content curation/filtering algorithms used by social media companies. This is because this thesis deals with opinion dynamics on social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
There are two major groups of opinion dynamics models, namely binary models and continuous approaches. These will be discussed in the following two sections. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.
% talk about opinion dynamics, different models, resistance, etc
\section{Binary models}

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent has an opinion, then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive such a system to converge to unanimity and does an opinion, that is, initially, a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan \cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}.

\section{Continuous models}

\section{Algorithmic personalization}

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies didn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that one can process \cite{Bozdag2013}.
% search for literature to back this up! (Bozdag2013, read this!!)
% talk about the algorithms proposed by perra and rocha

>>>>

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics.\\
The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals was well know, than the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. \\
However, the use of statistical physics may also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. Note that this is not only an important factor in social physics, but also in regular physical research and science in general.
% talk about history

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. This chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure. Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary models and continuous approaches. These will be discussed in the following two sections, Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn people is discussed and, in Section \ref{thesisOp}, a summary of the opinion dynamics model used in this thesis is given.

\section{Binary models}\label{Bin}
% Talk about Ising model???

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent has an opinion, then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive such a system to converge to unanimity and does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of imbalance and a notion of `minority' and `majority' opinion)\cite{Benjamini2016}. It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}.
% https://arxiv.org/pdf/0710.3256.pdf Contains a mean field mathimatical description, interesting?

\subsection{The probabilistic majority model}

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors has opinion B), user $i$ will updates his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

\subsection{The Voter model}
% https://arxiv.org/pdf/0710.3256.pdf

\section{Continuous models}\label{Con}
% https://arxiv.org/pdf/0710.3256.pdf, bounded confidence methods
% https://www.researchgate.net/publication/235737446_Mixing_Beliefs_Among_Interacting_Agents defuant model
% http://jasss.soc.surrey.ac.uk/5/3/2/2.pdf
% http://jasss.soc.surrey.ac.uk/9/3/8.html

\section{Algorithmic personalization}\label{filter}

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that one can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{Modelling opinion dynamics in the age of algorithmic personalization} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from its friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line; the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the users and then selects the first $Q$ elements
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC method is mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

\section{The activation mechanism}\label{actMech}

\section{The presence of stubborn people}\label{stubb}

\section{The opinion dynamics model used in this thesis}\label{thesisOp}

% https://arxiv.org/pdf/0710.3256.pdf Talk about agent-based models??
